---
title: "Assessment 0 - 01: Data and Introduction"
output: html_notebook
---

## The brief

The brief was to:


## The team


## Library requirements

```{r}
if (!require("fs")) install.packages("fs")
if (!require("googledrive")) install.packages("googledrive")
library("fs")
library("googledrive")
```


## The data

We each started by searching out a number of datasets. These varied greatly in context, volume and existing analysis. As such there are a number of datasets that should be downloaded. 

To obtain this dataset in a convenient format, we will download it and process it into a standard form (containing headers).

We place raw data in the `data` folder of our root. (Note: we are in the /report directory now). So our file system will look like this:

* /data
  * /data/raw
    * /data/raw/4SICS.csv
  * /data/processed
    * /data/raw/4SICS.tab



### Get the data
```{r}
destfile <- "D:/Adam/Bristol/Units/Year 4/Data Science Toolbox/Assessments/Asessment 0/DST---Assessment-0-Group-2/data/raw/4SICS.csv"
drive_download(as_id(""))

``` 


### Get the names for the data
```{r}
cnames=c('R1voltAngleA','R1voltAngleB','R1voltAngleC','R1voltMagA',
    'R1voltMagB','R1voltMagC','R1curAngleA','R1curAngleB','R1curAngleC',
    'R1curMagA','R1curMagB','R1curMagC','R1posVoltAngle',
    'R1negVoltAngle','R1zeroVoltAngle','R1posVoltMag','R1negVoltMag',
    'R1zeroVoltMag','R1posCurAngle','R1negCurAngle','R1zeroCurAngle',
    'R1posCurMag','R1negCurMag','R1zeroCurMag','R1relayFreq','R1relayDeltaFreq',
    'R1relayImp','R1relayImpAngle','R1status','R2voltAngleA','R2voltAngleB','R2voltAngleC','R2voltMagA',
    'R2voltMagB','R2voltMagC','R2curAngleA','R2curAngleB','R2curAngleC',
    'R2curMagA','R2curMagB','R2curMagC','R2posVoltAngle',
    'R2negVoltAngle','R2zeroVoltAngle','R2posVoltMag','R2negVoltMag',
    'R2zeroVoltMag','R2posCurAngle','R2negCurAngle','R2zeroCurAngle',
    'R2posCurMag','R2negCurMag','R2zeroCurMag','R2relayFreq','R2relayDeltaFreq',
    'R2relayImp','R2relayImpAngle','R2status','R3voltAngleA','R3voltAngleB','R3voltAngleC','R3voltMagA',
    'R3voltMagB','R3voltMagC','R3curAngleA','R3curAngleB','R3curAngleC',
    'R3curMagA','R3curMagB','R3curMagC','R3posVoltAngle',
    'R3negVoltAngle','R3zeroVoltAngle','R3posVoltMag','R3negVoltMag',
    'R3zeroVoltMag','R3posCurAngle','R3negCurAngle','R3zeroCurAngle',
    'R3posCurMag','R3negCurMag','R3zeroCurMag','R3relayFreq','R3relayDeltaFreq',
    'R3relayImp','R3relayImpAngle','R3status','R4voltAngleA','R4voltAngleB','R4voltAngleC','R4voltMagA',
    'R4voltMagB','R4voltMagC','R4curAngleA','R4curAngleB','R4curAngleC',
    'R4curMagA','R4curMagB','R4curMagC','R4posVoltAngle',
    'R4negVoltAngle','R4zeroVoltAngle','R4posVoltMag','R4negVoltMag',
    'R4zeroVoltMag','R4posCurAngle','R4negCurAngle','R4zeroCurAngle',
    'R4posCurMag','R4negCurMag','R4zeroCurMag','R4relayFreq','R4relayDeltaFreq',
    'R4relayImp','R4relayImpAngle','R4status','contPanLog1','contPanLog2',
    'contPanLog3','contPanLog4','relayLog1','relayLog2','relayLog3',
    'relayLog4','snortLog1','snortLog2','snortLog3','snortLog4',
    'marker')
write.table(cnames,
            file=path_wd("..","data","raw","Morris1.names"),
            row.names=FALSE,
            col.names=FALSE,
            quote=FALSE
            )
```

```{r}
cnames=c('Packet Number','Time','Source IP','Destination IP',
    'Protocol','Packet Length','Info')
write.table(cnames,
            file=path_wd("..","data","raw","4SICS.names"),
            row.names=FALSE,
            col.names=FALSE,
            quote=FALSE
            )
```

### Create the processed data output.

```{r}
DF=read.csv(path_wd("..","data","raw","data1.csv"),header=TRUE)
for (n in c(2:15)){
  fn <- paste("data",n,".csv",sep="")
  path_wd("..","data","raw",fn)
  df=read.csv(path_wd("..","data","raw",fn),header=TRUE)
  DF=rbind(DF,df)
}
colnames(DF)=cnames
write.csv(DF,
            path_wd("..","data","processed","Morris1.csv"),
            quote=FALSE,row.names=FALSE)
```

```{r}
DF=read.csv(path_wd("..","data","raw","4SICS.csv"),header=TRUE)
colnames(DF)=cnames
write.csv(DF,
            path_wd("..","data","processed","4SICS.csv"),
            quote=FALSE,row.names=FALSE)
```

### Summary:

