---
title: "Assessment 0 - 01: Data and Introduction"
output: html_notebook
---

## The brief

The brief was to:


## The team


## Library requirements

```{r}
if (!require("fs")) install.packages("fs")
library("fs")
```


## The data

We each started by searching out a number of datasets. These varied greatly in context, volume and existing analysis. As such there are a number of datasets that should be downloaded. 

To obtain this dataset in a convenient format, we will download it and process it into a standard form (containing headers).

We place raw data in the `data` folder of our root. (Note: we are in the /report directory now). So our file system will look like this:

* /data
  * /data/raw
    * /data/raw/4SICS.csv
  * /data/processed
    * /data/raw/4SICS.tab



### Get the data
```{r}
download.file("https://drive.google.com/file/d/14O_LdaJ-2lT7BN5pVjaFrtymk8L_tzRF/view?usp=sharing",path_wd("..","data","raw","4SICS.csv"))
```

### Get the names for the data


### Create the processed data output.


### Summary:

