---
title: "Assessment 0 - 01: Introduction"
output: html_notebook
---

## The brief

The brief was to:

* Give us an introductory experience to future assessment environments
* Enable us to explore working in collaboration with peers
* Provide us with opportunity to gain further programming experience.

Through the process of analysing cyber security datasets. To do this we gathered a number of resources together in the form of both datasets and methods to analyse these datasets. We each chose a datset that we thought would be interesting to analyse, as we believed that it would be better to look into more datasets rather than focusing on just one. 

## The team
The team consisted of Adam, Bill and Oliver. Adam had experience with Git, R and Python, so took charge of setting up the Github repository. He chose to look at ICS datasets because he thinks they provide an interesting opportunity for ML algorithms. Bill had used a little R but was new to Git. He chose to look at a dataset of URLs, some of which were known phishing sites. The aim of Bills analysis was to compare the benign and phishing URLs, and investigate what derivable features could help with classification.

and Oliver had experience in ...

## Library requirements

```{r}
if (!require("fs")) install.packages("fs")
if (!require("googledrive")) install.packages("googledrive")
library("fs")
library("googledrive")
```


## The data

We each started by searching out a number of datasets. These varied greatly in context, volume and existing analysis. As such there are a number of datasets that should be downloaded. 

To obtain this dataset in a convenient format, we will download it and, if necessary, process it into a standard form.

We place data in the raw or processed folder depending on the stage of processing, both of which are in the `data` folder of our root. So our file system will look like this:

* /data
  * /data/processed
    * /data/processed/4SICS.csv
    * /data/processed/Morris1.csv



### Get the data
```{r}
drive_deauth()
drive_download(as_id("14O_LdaJ-2lT7BN5pVjaFrtymk8L_tzRF"), path = "../data/processed/4SICS.csv", overwrite = TRUE)
drive_download(as_id("1KD1MiHkJObkwE95mimI19q4SkxfZCHcy"), path = "../data/processed/Morris1.csv", overwrite = TRUE)
``` 

